How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-07-24 doc/response_review.tex

\begin{refsection}

\maketitle

\linenumbers{}
\modulolinenumbers[3]

Below are our responses to the editor and reviewer comments that we received on our original manuscript. We would like to thank the editor and the three reviewers for their thoughtful and insightful comments which have greatly improved the manuscript.

@@

\subsection{Reviewer 1}

\begin{bfseries}
  Smits and Finnegan use the Neptune database to check which combination of potentially influencing variables has the best predictive power of extinction risk in the marine plankton. They use sophisticated logistic modelling of the time-binned raw data to suggest that considering changes of parameters only marginally improves prediction accuracy. Overall, this is a very well written, although difficult to comprehend, account, which is to the point and rich in insights.

  While the main result is well founded by the sophisticated analyses (in fact too sophisticated to be understandable in all details by the reviewer), it is somewhat disappointing, as the relative importance of parameters in multivariate models remain vague. It is also disappointing to see that change of range had a negligible effect in the multivariate context, as this is one of the main arguments for using the fossil record in extinction risk assessments (besides taxon age and intrinsic turnover rates). I argue that, to really discard change of range, the authors would need to look at longer trajectories of range. Bin-to-bin changes at this temporal resolution (1 myr) are perhaps too noisy to inform models. I am not asking for additional analyses for this paper, but this point should be added to the discussion.
\end{bfseries}

Thank you for this suggestion; see above re. the change we made in the timespace of historical covariates.

@@

Thank you for pointing out this issue in our text and we apologize for the lack of clarity. We have updated our methods section to further explain cross-validation and how we have implicitly dealt with model complexity because of the linked definitions of cross-validation and AIC. From our updated text:

``Cross-validation is a procedure for estimating a models expected out-of-sample error. Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}. A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.''

\begin{bfseries}
  The methods are very well explained (especially in the supplement), although uncertainties in parameter estimates merit some further prose. Using maximum great circle distance (GCD) as a measure of geographic range is ok but prone to sampling bias. Better sampled intervals will inevitably have greater mean ranges than more poorly sampled intervals. While the scaling applied by the authors takes different means and variance into account it may not be sufficient. If GCD is used, I’d recommend normalizing to the maximum observed geographic range in an interval prior to scaling. Grid cell occupancy or minimum spanning trees are alternative measures, which are worthy of exploration. 
\end{bfseries}

Thank you both for the suggestion to move to minimum spanning tree measure of geographic range. We have updated our analysis using this new range metric. The change in range metric did not substantively change our interpretation of the results.

@@

\begin{bfseries}
  The supplement is good but lacks sensitivity tests and a rationale of why the specific parameters and not others where chosen in the models. 
\end{bfseries}

Thank you for this suggestion. We have improved the language in our methods section surrounding in-sample verus out-of-sample forecasts. Our in-sample forecasts are a type of sensitivity test.

From the updated manuscript:

Line 173: ``In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to correctly classify the data to which it was fit. Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data. If our models have poor in-sample forecasting performance, than our models are not adequate descriptors of the data and will most likely make poor out-of-sample predictions. In-sample forecasting measures, however, are not the same as understanding our models' ability to forecast future extinctions or if our models are overfit to our data and produce biased out-of-sample estimates \citep{ESL}.''

@@

And:

Line 232: ``The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis. Comparison between the posterior distributions \dots''

@@

  I only have two broader thoughts and two more specific notes:

  I find it interesting that the predictability of extinction doesn’t obviously differ between background intervals and intervals with more unusual environmental perturbations (such as the PETM or E-O transition). Perhaps that’s because those events weren’t mass extinctions in the vein of the P/T or K/Pg. There’s also a fair amount of volatility in the AUC time-series. Nevertheless, does this consistent predictive ability regardless of differences in conditions say anything interesting about extinction predictions?

  The caveat that human impacts may dramatically alter extinction risk seems like an important one and, as it stands, seems like it could undercut the (rather terse) final conclusion. Would conservation decisions really be bolstered by including fossil data? I would agree with that (and of course I would, as a paleobiologist), but it might be useful to expand a bit on this topic. Do you mean that a model such as yours, incorporating geographic range and other such parameters, would be helpful? In what way could it be used? Or do you mean using models based on past intervals with environmental changes inferred to be similar – the PETM as an analogue for warming and acidification, for example? I wonder if it would be helpful to draw a stronger connection linking your finding to conservation biology, given the different stressors now.
\end{bfseries}

Thank you for your suggestion to expand upon the implciations of the human impact on extinction risk and how that might affect how well we can predict future extinction events. We have added a section to our discussion in an attempt to address this.

Line 318: ``An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}. The CP model assumes that the effects of the historical covariates are constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes. Thus, it might be preferable to use a model similar to our V model which allows extinction risk and selectivity to vary over time. For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.''

On a related note, it is notable that there are no obvious consistent changes in average model performance during episodes of climatic environmental change such as the mid-Eocene to early Oligocene (Fig. 1b), although this interval is characterized by elevated extinction among several planktonic groups \citep{Prothero1994,Wade2008,Kamikuri2012}. This suggests that relationships between the simple covariates included in our models and extinction risk were relatively stable though this interval.  However, it is likely that more complex models accounting for other aspects of ecology and biogeography might exhibit reduced forecasting performance during this and other climate change episodes.

@@

  The authors need some discussion and justification on this.
\end{bfseries}

We have added a section to the Methods discussing this data limitation

From the updated manuscript:

Line 134: ``Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature. The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit. Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic. Unfortunately, that type of data does not exist. So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.''

### 2019-07-24 doc/article.tex
repsonse reviews. second round started.

Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}. We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic foram and seawater Mg/Ca isotope ratio information. Because seawater Mg/Ca ratio has changed over time, the method to estimate temperature used \citet{Cramer2011} attempts to account for this unknown in order to obtain accurate, albeit uncertain, temperature estimates. Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My interval between 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

@@

Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature. The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit. Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic. Unfortunately, that type of data does not exist. So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.

@@

Cross-validation is a procedure for estimating a model's expected out-of-sample error. Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}. A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.

@@

@@

The primary focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates. A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section S2).

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation. The area under the receiver operating charactered curve (AUC) was calculated for each model. these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively. Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in the dataset. Models with higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
 \label{fig:auc_compare}
\end{figure}

@@

The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis. Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that the parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, while there is substantial overlap in the posterior distributions of the forecasts from the other three models (Fig. \ref{fig:auc_hist}).

@@

In-sample forecasts from the four models over time are broadly similar between taxonomic groups (Fig. \ref{fig:auc_taxon_time}). In-sample forecasts for diatoms are the weakest of the four taxonomic groups as all four models have several intervals with no predictive power (AUC not significantly greater than 0.5). The best in-sample forecast results are for radiolarians, for which all models have at most 1 interval with little predictive power. In-sample forecasts over time by taxonomic group are broadly consistent between the four models.

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
 \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:auc_taxon_time}
\end{figure}

@@

We estimated expected out-of-sample forecasting performance using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. Out-of-sample forecast AUC estimates exhibit a broader range than in-sample estimates, with AUC ranging between approximately 0.7 and 0.85 (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). While the VP model has the best in-sample forecasting performance (Fig. \ref{fig:auc_hist}) this model performs poorly at out-of-sample forecasting compared to the other models (Fig. \ref{fig:fold_auc}). The poor out-of-sample performance suggests that this most complex model is overfit and that one of the simpler models would be preferable for predicting future extinctions. Thus the models that include both historical covariates (e.g. change in geographic range) and time-varying effects produce biased extinction forecasts. Interestingly, models that include either historical covariates but assume constant effects (CP) or do not include historical covariates but include time-varying effects (V) perform similarly when forecasting future extinction events (Fig. \ref{fig:auc_compare}).

As noted above there were some time intervals in which in-sample forecasts were no better than random (Fig. \ref{fig:auc_taxon_time}). Such intervals are generally much rarer for out-of-sample forecasts. The major exception to this pattern are diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of the four models' ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:fold_auc_taxon_time}
\end{figure}

We compared the difference in AUC estimates from the out-of-sample forecasts to the AUC estimates from in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}); a difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that out-of-sample forecasts actually outperform in-sample forecasts, whereas negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between out-of-sample and in-sample forecasts are rare and tend not to cluster in time, consistent with the broad visual congruence between the in-sample and out-of-sample performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
 \caption{Comparing out-of-sample and in-sample forecasts. Congruence between in-sample and out-of-sample indicates that a model is not necessarily overfit to the data. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that the out-of-sample forecasts are actually higher performing than the in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:oos_ins_diff}
\end{figure}

@@

The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the Foraminifera may be informative.

An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}. The CP model assumes that the effects of the historical covariates are constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes. Thus, it might be preferable to use a model similar to our V model which allows extinction risk and selectivity to vary over time. For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.

>>>>

Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}. We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic Foraminifera and seawater Mg/Ca isotope ratio information. Because seawater Mg/Ca ratio has changed over time, the method to estimate temperature used \citet{Cramer2011} attempts to account for this unknown in order to obtain accurate, albeit uncertain, temperature estimates. Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My interval between 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

@@

Mg/Ca based temperature estimates are measured from benthic Foraminifera, and are an estimate of deep water ocean temperature. The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit. Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic. Unfortunately, that type of data does not exist. So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.

@@

Cross-validation is a procedure for estimating a model's expected out-of-sample error. Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}. A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces an entire posterior distribution of estimates.

@@

To reiterate, the primary focus of this study is on understanding how well our models forecast future extinction events by comparing our in-sample and out-of-sample forecast estimates. A presentation of the posterior estimates for the regression coefficient estimates from our VP model (Table \ref{tab:model_def}) is available our Supplemental Materials (Section S2).

@@

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
   %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{auc_hist_full.png}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
   %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{fold_auc_full.png}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation. The area under the receiver operating charactered curve (AUC) was calculated for each model. these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively. Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in the dataset. Models with higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
 \label{fig:auc_compare}
\end{figure}

@@

Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that the parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, while there is substantial overlap in the posterior distributions of the forecasts from the other three models (Fig. \ref{fig:auc_hist}).

@@

In-sample forecasts from the four models over time are broadly similar among taxonomic groups (Fig. \ref{fig:auc_taxon_time}). In-sample forecasts for diatoms are the weakest of the four taxonomic groups as all four models have several intervals with no predictive power (AUC not significantly greater than 0.5). The best in-sample forecast results are for radiolarians, for which all models have at most 1 interval with little predictive power. The pattern of high and low in-sample forecast performance is broadly similar among the four models.

\begin{figure}[ht]
 \centering
 %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{auc_taxon_time_full.png}
 \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:auc_taxon_time}
\end{figure}

@@

Out-of-sample forecast AUC estimates, based on five-fold cross validation \citep{Arlot2010,Bergmeir2016}, exhibit a broader range than in-sample estimates, with AUC ranging between approximately 0.7 and 0.85 (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). While the VP model has the best in-sample forecasting performance (Fig. \ref{fig:auc_hist}) this model performs poorly at out-of-sample forecasting compared to the other models (Fig. \ref{fig:fold_auc}). The poor out-of-sample performance suggests that this most complex model is overfit and that one of the simpler models would be preferable for predicting future extinctions. Thus the models that include both historical covariates (e.g. change in geographic range) and time-varying effects produce biased extinction forecasts. Interestingly, models that include either historical covariates but assume constant effects (CP) or do not include historical covariates but include time-varying effects (V) perform similarly when forecasting future extinction events (Fig. \ref{fig:auc_compare}).

As noted above, there were some time intervals in which in-sample forecasts were no better than random (Fig. \ref{fig:auc_taxon_time}). Such intervals are generally much rarer for out-of-sample forecasts. The major exception to this pattern are diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{fold_auc_taxon_time_full.png}
 \caption{Comparison of the four models' ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:fold_auc_taxon_time}
\end{figure}

We compared the difference in AUC estimates from the out-of-sample forecasts to the AUC estimates from in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}); a difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that out-of-sample forecasts actually outperform in-sample forecasts, whereas negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between out-of-sample and in-sample forecasts are rare and tend not to cluster in time, consistent with the broad visual congruence between the in-sample and out-of-sample performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). One of the few multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model between 35 Mya and approximately 28 Mya (Fig. \ref{fig:oos_ins_diff}). There exist similar periods of worse out-of-sample forecasting performance for other combinations of taxonomic group and predictive model, but none are as long as that for Radiolaria.

\begin{figure}[ht]
 \centering
 %\includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{auc_diff.png}
 \caption{Comparing out-of-sample and in-sample forecasts. Congruence between in-sample and out-of-sample indicates that a model is not necessarily overfit to the data. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that the out-of-sample forecasts are actually higher performing than the in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:oos_ins_diff}
\end{figure}

@@

The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many Foraminifera species \citep{Ezard2011}. However, comparable information does not exist for all Foraminifera species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the Foraminifera may be informative.

An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}. The CP model assumes that the effects of the historical covariates are constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes. Thus, it might be preferable to use a model similar to our V model which allows extinction risk and selectivity to vary over time. For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.

### 2019-07-24 doc/letter_editor.tex
repsonse reviews. second round started.

Please find enclosed our manuscript entitled ``How predictable is extinction? Forecasting species survival at million-year timescales'' by myself (Peter D. Smits) and Seth Finnegan. In this study we asked a relatively simple yet fundamental question for the emerging field of conservation paleobiology: how well do models conditioned on past observations predict future extinction events? To answer this question we analyzed the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examined how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state.

We used cross-validation to estimate the expected forecast performance of our model for future observations. Our models were found to have a 70-80\% probability of correctly forecasting the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction risk have varied only modestly through time. Including specific information about individual species trajectories resulted in only extremely marginal improvement in fit, suggesting that, at least at the million-year timescale, extinction risk is best described by dominantly Markovian (e.g. memoryless) models.

The relative quality and consistency of our models’ out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. The results of this simple exercise bolster the case that fossil data can meaningfully inform present and future conservation decisions. We therefore believe that our study will be of general interest to paleontologists, conservation biologists, ecologists, and evolutionary biologists.  We look forward to your decision. Please send all correspondence regarding this manuscript to me via my email address (psmits@berkeley.edu) or that of my co-author, Seth Finnegan (sethf@berkeley.edu).

@@

  \encl{Article, supplementary text}.

>>>>

Please find enclosed our revised manuscript entitled ``How predictable is extinction? Forecasting species survival at million-year timescales'' by myself (Peter D. Smits) and Seth Finnegan. We would like to thank the editor and the three reviewers for their thoughtful comments which greatly improved our manuscript.

In this study we asked a relatively simple yet fundamental question for the emerging field of conservation paleobiology: how well do models conditioned on past observations predict future extinction events? To answer this question we analyzed the well-sampled fossil record of Cenozoic planktonic microfossil taxa. We examined how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Our models were found to have a 70-80\% probability of correctly forecasting the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction risk have varied only modestly through time.

The relative consistency of our models’ out-of-sample forecasting performance is encouraging given that these estimates are based on limited biological and environmental information. The results of this simple exercise bolster the case that fossil data can meaningfully inform present and future conservation decisions. We therefore believe that our study will be of general interest to paleontologists, conservation biologists, ecologists, and evolutionary biologists.  We look forward to your decision. Please send all correspondence regarding this manuscript to me via my email address (psmits@berkeley.edu) or that of my co-author, Seth Finnegan (sethf@berkeley.edu).

@@

  \encl{Article, supplementary text, response to reviewers}.
